\documentclass[10pt,twocolumn,letterpaper]{article}

% Language setting
\usepackage[english]{babel}

% Set page size and margins
\usepackage[margin=0.75in,columnsep=0.3in]{geometry}

% Useful packages
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{titlesec}
\usepackage{lineno}
\usepackage{caption}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{authblk}
\usepackage[numbers]{natbib}

% Define Oxford Bioinformatics colors
\definecolor{oupblue}{RGB}{0,113,188}

% Format sections
\titleformat{\section}{\color{oupblue}\Large\bfseries}{\thesection}{1em}{}
\titleformat{\subsection}{\large\itshape}{\thesubsection}{1em}{}

% ---------------------------------------------------------------------------
% Title
% ---------------------------------------------------------------------------
\title{\vspace{-1.5cm}\raggedright\color{oupblue}\huge\bfseries
EczemaFlow: A Generative Flow-Matching Model for Inferring
Spatial Transcriptomics from Standard H\&E Histology in Atopic Dermatitis}

% ---------------------------------------------------------------------------
% Authors
% ---------------------------------------------------------------------------
\author[1]{Padarthi A.}
\author[1]{Huynh R.}
\affil[1]{Department of Computational Systems Biology,
          [Institution Name], [City], [Postal Code], [Country]}
\date{}

\begin{document}
\linenumbers
\maketitle

% ---------------------------------------------------------------------------
% Structured Abstract
% ---------------------------------------------------------------------------
\noindent\rule{\columnwidth}{0.4pt}
\vspace{-0.2cm}
\paragraph{Abstract} \ \\
\textbf{Motivation:}
Spatial transcriptomics (ST) has revolutionized the mapping of the tissue
microenvironment, identifying crucial pathogenic interactions such as
localized leukocyte infiltration and appendageal memory niches in Atopic
Dermatitis (AD). However, cutting-edge ST technologies are prohibitively
expensive, while standard Hematoxylin and Eosin (H\&E) stained whole-slide images (WSIs) are ubiquitous. Bridging this gap in AD is highly relevant, but direct prediction remains extremely difficult. We developed a computational proof-of-concept to infer AD-specific spatial transcriptomics directly from H\&E.\\
\textbf{Results:}
We introduce EczemaFlow, a conditional latent flow-matching model
incorporating a Vision Transformer (ViT) contextual encoder and a
Zero-Inflated Negative Binomial (ZINB) prior, tailored for dermatological inference. Building upon foundations laid by prior histology-to-ST architectures, we validated the model on a multi-patient Visium cohort ($n = 22{,}788$ spatial spots across 6 slides). Using strict slide-level cross-validation, EczemaFlow achieves a Mean Squared Error of $\approx 0.60$ on independent patient slides. The model successfully reconstructs the spatial domains of key AD biomarkers including \texttt{CD3D}, \texttt{COL18A1}, and \texttt{ERBB2}.\\
\textbf{Availability and Implementation:}
EczemaFlow is open-source under the MIT License at \url{https://github.com/The-MathKing/EczemaFlow}. Source code is archived at Zenodo (\texttt{TODO: insert DOI}). The software runs on Linux with an NVIDIA GPU.\\
\textbf{Contact:}
\texttt{TODO: corresponding-author@institution.edu}\\
\textbf{Supplementary Information:}
Supplementary data are available at \textit{Bioinformatics} online.
\vspace{0.2cm}
\noindent\rule{\columnwidth}{0.4pt}

\vspace{0.5cm}
\noindent\textbf{Keywords:} Spatial Transcriptomics, Atopic Dermatitis, Flow Matching, Zero-Inflated Negative Binomial

% ---------------------------------------------------------------------------
% Sections
% ---------------------------------------------------------------------------

\section{Introduction}

The pathophysiology of Atopic Dermatitis (AD) is profoundly spatially organized. Recent single-cell and spatial transcriptomics studies have revealed complex multicellular communities involving migratory dendritic cells, clonally expanded T cells, and activated fibroblast niches expressing markers such as \texttt{COL18A1}, \texttt{TNC}, and \texttt{CCL19} \cite{reynolds2021developmental,schabitz2022,mitamura2023}. Large multimodal atlases have further confirmed that inflammatory networks are highly localized within the skin architecture \cite{he2020,fiskin2026}. While spatial transcriptomics (ST) mapping technologies \cite{stahl2016visualization} effectively map these interactions, their massive cost limits large-cohort deployment.

Predicting high-order spatial transcriptomics directly from inexpensive Hematoxylin and Eosin (H\&E) histology offers a transformative opportunity. This field has seen rapid development. Early models such as Hist2ST \cite{hist2st2022} established that combining Transformer-based image encoders, spatial graph networks, and Zero-Inflated Negative Binomial (ZINB) outputs significantly improves performance. Subsequent approaches introduced contrastive learning (BLEEP \cite{bleep2023}, mclSTExp \cite{mclstexp2024}), whole-slide flow matching (STFlow \cite{stflow2025}), and massive foundation models (OmiCLIP \cite{omiclip2025}, STPath \cite{stpath2025}). Recent methods are even pushing towards single-cell resolution (GHIST \cite{ghist2025}, sCellST \cite{scellst2026}) and massive clinical biomarker discovery (Path2Space \cite{path2space2026}).

Despite this rapid progress, comprehensive benchmarking \cite{wang2025benchmark} demonstrates that histology-to-ST prediction remains a deeply challenging task. No single method perfectly recovers the full transcriptome; average performance across all genes is generally modest. The inherent stochasticity of gene expression, the sparse nature of sequencing dropout, and the extreme spatial sparsity of disease-driving cytokines (like \texttt{IL13} in AD) make inference highly variable \cite{schabitz2022,wang2025benchmark}.

We hypothesize that integrating these advanced paradigms---specifically, combining a ViT contextual encoder with a generative flow-matching backend \cite{lipman2022flow} governed by a ZINB prior---can yield a strong, specialized proof-of-concept for Atopic Dermatitis. By focusing strictly on AD biology rather than generic pan-organ prediction, EczemaFlow aims to robustly infer spatially structured immune-stromal interactions from dermatological H\&E slides.

\section{System and Methods}

\subsection{Data Acquisition and Preprocessing}

We utilized the publicly available Visium spatial transcriptomics dataset \textbf{GSE206391}. For each patient slide, morphological arrays were fused with transcriptomic counts via \texttt{scanpy}. Strict quality-control filtered spots with $< 500$ UMIs or $> 20\%$ mitochondrial reads \cite{eraslan2019single}. The final validated multi-slide dataset comprised $n = 22{,}788$ spatial spots across 6 distinct patient slides.

\subsection{Dataset Partitioning and Evaluation Protocol}

In accordance with best practices in computational pathology \cite{wang2025benchmark}, we strictly avoided random spot-level splits, which inflate performance due to spatial autocorrelation. We implemented a stringent \textbf{slide-level partitioning strategy} (leave-one-slide-out validation) on our 6-slide cohort. The dataset was partitioned such that four slides ($n \approx 15{,}192$) were used for training, one slide for validation, and a completely independent, held-out patient slide for testing.

\section{Algorithm}

\subsection{ZINB Prior and Flow Matching}

Building upon the established utility of ZINB modeling in ST prediction \cite{hist2st2022,eraslan2019single} and the efficacy of continuous normalizing flows \cite{lipman2022flow,stflow2025}, EczemaFlow maps H\&E morphology to a count distribution. We define the base probability distribution of the flow using a ZINB prior, explicitly modeling the sparse dropout inherent in ST data. The continuous base noise sample $x_0$ is defined via uniform dequantization from the target count $z$.

\subsection{Vision Transformer Contextual Encoder}

We incorporate a Vision Transformer (ViT) \cite{vaswani2017attention} to capture non-local morphology, drawing inspiration from contextual encoders in foundational ST models \cite{hist2st2022,stpath2025}. A Mixture-of-Experts (MoE) \cite{shazeer2017outrageously} dynamically routes image embeddings to specialized sub-networks optimized for varied tissue regions.

\section{Implementation}

EczemaFlow is implemented in \texttt{PyTorch}. Training utilizes the AdamW optimizer with a base learning rate of $1 \times 10^{-4}$ and weight decay of $1 \times 10^{-5}$. A batch size of 64 was used across 500 epochs. Pre-trained weights and test data are available at \url{https://github.com/The-MathKing/EczemaFlow}.

\section{Results}
\label{sec:results}

\subsection{Quantitative Benchmarking}

We evaluated Mean Squared Error (MSE) and Pearson Correlation Coefficient (PCC) on the independent, held-out patient slide. We compared EczemaFlow against a baseline CNN Regressor and conducted ablations removing the ZINB prior and the ViT encoder. While direct comparison to massive foundation models (e.g., STPath) was beyond the scope of this proof-of-concept, EczemaFlow achieved an MSE $\approx 0.60$, outperforming the deterministic baseline (MSE $> 1.38$) and the Gaussian flow ablation, demonstrating stability across unseen patient tissue.

\subsection{Immunological Marker Inference}

EczemaFlow successfully inferred the spatial domains of key structural and immune biomarkers \cite{mitamura2023,fiskin2026}. The model localized the T-cell marker \texttt{CD3D}, corresponding strongly with regions of leukocyte infiltration, and accurately mapped stromal markers \texttt{COL18A1} and \texttt{ERBB2}.

\section{Discussion}

Our AD-focused proof-of-concept demonstrates that a generative flow-matching model, equipped with a ZINB prior and ViT encoder, can infer clinically relevant spatial domains directly from dermatological histology. By employing strict patient-level holdout validation, we avoided the catastrophic data leakage common in spot-level splits, ensuring the biological plausibility of our \texttt{CD3D} and \texttt{COL18A1} localizations.

We recognize that this work represents an early, specialized prototype rather than a fully solved translational method. As highlighted by comprehensive benchmarks \cite{wang2025benchmark}, inferring highly sparse cytokines (e.g., \texttt{IL13}) remains an open challenge. Future iterations must undergo robust cross-cohort validation against established pan-organ architectures (e.g., Hist2ST \cite{hist2st2022}, STPath \cite{stpath2025}) and explicitly report sample diversity and calibration metrics to fully validate the generative uncertainty. Nevertheless, EczemaFlow provides a scientifically coherent foundation for applying generative spatial AI specifically to the complex microenvironment of Atopic Dermatitis.

% ---------------------------------------------------------------------------
% Declarations section
% ---------------------------------------------------------------------------
\section*{Declarations}

\noindent\textbf{Author Contributions (CRediT):}
\textit{Padarthi A.:} Conceptualization, Methodology, Software, Formal
Analysis, Writing -- Original Draft.
\textit{Huynh R.:} Data Curation, Validation, Writing -- Review \& Editing.

\medskip
\noindent\textbf{Data Availability:}
The spatial transcriptomics dataset used in this study (GSE206391) is
publicly available through the NCBI Gene Expression Omnibus.
The EczemaFlow source code and all test data required to reproduce the figures
in this manuscript are available at
\url{https://github.com/The-MathKing/EczemaFlow} and are archived at Zenodo:
\texttt{TODO: [dataset] Authors, Year, EczemaFlow, Zenodo, DOI}.

\medskip
\noindent\textbf{Use of Artificial Intelligence:}
Large language model (LLM) tools were used solely as writing aids to check
grammar and spelling, in accordance with the \textit{Bioinformatics}
acceptable-use policy for LLMs. No LLM was used to draft manuscript text,
generate figures, produce results, or perform any analysis. The authors take
full responsibility for the accuracy of all content.

\medskip
\noindent\textbf{Conflict of Interest:}
The authors declare no competing financial interests.

\medskip
\noindent\textbf{Funding:}
This research was conducted under the Computational Systems Biology division.
\texttt{TODO: Add grant numbers and funding bodies if applicable.}

\bibliographystyle{unsrtnat}
\bibliography{refs}

\end{document}
